Supplementary Information: Drought sensitivity is not a species trait: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-11-01 doc/response_to_reviewer_R2.Rmd

---
title: 'Responses to the Reviewer'
output: 
    bookdown::word_document2: 
    #   reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
      reference_docx: word-styles-reference-singlespace.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document:
    # engine: xelatex 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
# csl: new-phytologist.csl
csl: ecology-letters.csl 
suppress-bibliography: TRUE
---

Dear Editor,

We are pleased to submit a revised version of our manuscript, "Drought response is not a species trait: tropical tree drought sensitivity is shaped by drought characteristics, species adaptations, and individual microenvironments" (ELE-00826-2025) for consideration for publication in Ecology Letters. 
We have addressed all points raised by the Data Editor as detailed in our responses to the reviewers and have updated the files provided in the Zenodo library to ensure reproducibility of our results. 
We have also updated the Data Availability Statement in the manuscript to reflect this.

Thank you for considering this revised version for publication. 
We look forward to your response.

Sincerely,
On behalf of all coauthors,  
Krishna Anujan

\newpage

## Referees' comments to the author(s):

### Referee: 1

Comments for the Authors
Congratulations on sharing the code. However, the data are only shared upon request and are not open, contrary to the request of Ecology Letters, so the analyses are not reproducible. Briefly, the README is present and describes all the content. The scripts and manuscript codes are made available. The only thing lacking is metadata to understand the data's content. The only necessary step for reproducibility would be to share the data. To improve upon this, the author could consider defining package versions, or even better, using an environment manager such as renv. They could also take advantage of a forge such as GitHub to make things easier and improve reproducibility. Consequently, I rated the manuscript as '1) Inadequate (data or code not given or insufficient)' as the data are not provided.

**RESPONSE: Thanks for the favourable review of our manuscript code. We have now updated the Zenodo repository [doi: 10.5281/zenodo.15777974](https://zenodo.org/records/15777975) to include the datasets required to reproduce these analyses as csvs. We have also included metadata csvs describing the columns in the data csvs. Finally, as suggested here by the Data Editor, we have also included a session_info.txt file that describes the hardware and software versions (including packages) used to run these analyses. We believe that this revision allows for full reproducibility of our results and figures. Moreover, the entire analyses was managed through a GitHub repository - accessible through the Zenodo link - so the versions are traceable and reproducible.**

### 2025-11-09 doc/manuscript_hkk_drought_sensitivity.Rmd
data availability statement updated
updated manuscript Rmd with the public datasets

<!-- **Title 1:** Sensitivity of tree growth to drought in a seasonally dry tropical forest using long-term dendrometer band measurements

**Title 2:** Contrasting effects of species and environmental predictors on tree growth sensitivity to different/two droughts in a seasonally dry tropical forest

**Title 3:** Diversity of species- and environment-mediated drought responses in a seasonally dry tropical forest

**Title 4:** Long-term dendrometer band measurements reveal diverse species- and environment-mediated drought responses in a seasonally dry tropical forest

**Title 5:** Deciduousness may exacerbate/worsen/amplify tree growth sensitivity during wet season droughts in seasonally dry tropical forests -->

@@

# Data Accessibility Statement {-}
<!-- This manuscript does not use novel code.  -->
<!-- The dendrometer band data used for this manuscript is archived in the ForestGEO data repository and will be made available upon request.  -->
<!-- The climate data used in this work are publicly available.  -->
Data and code required to reproduce the results and figures are archived in Zenodo with doi:10.5281/zenodo.15777974.

@@

```{r data summaries, echo=F, warning=F, message=F}
library(tidyverse)
tree.time <- readRDS("../data/HKK-dendro/sensitivity_data_formodels.RData")
# tree.time <- readRDS("data/HKK-dendro/sensitivity_data_formodels.RData")

tree.time.sp <- tree.time %>%
  group_by(spfull) %>%
  dplyr::summarise( # change to mm
    inc_annual_med = round(median(inc_annual * 10, na.rm = T), 2),
    inc_annual_sd = round(sd(inc_annual * 10, na.rm = T), 2)
  ) %>%
  arrange(inc_annual_med)

sens.sp <- tree.time %>%
  filter(yr %in% c(2010, 2015, 2020)) %>%
  group_by(spfull, yr) %>%
  dplyr::summarise(
    sens_med = round(median(sens.prop, na.rm = T), 2),
    sens_mean = round(mean(sens.prop, na.rm = T), 2),
    sens_sd = round(sd(sens.prop, na.rm = T), 2)
  ) %>%
  arrange(sens_med)

# spread this for each year

@@

Drought years were associated with lower plot-level growth and negative tree-level sensitivities, with the strongest negative response in 2015 (Fig `r fig_drought_responses`). 
Median growth rate across all trees and years was `r round(median(tree.time$inc_annual*10, na.rm=T),2)` mm (range `r round(min(tree.time$inc_annual*10, na.rm=T),2)`, `r round(max(tree.time$inc_annual*10, na.rm=T),2)`). 
The drought year 2015 had the lowest median increment `r round(median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2015]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% lower than the median across all years. 
In the 2010 and 2020 droughts, the median increments were `r round(median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T),2)` mm and `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm respectively, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2010]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% and `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),0)`% lower than the median across all years. 
<!-- In the 2020 drought, the median annual increment was `r round(median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T),2)` mm, `r round(-100*((median(tree.time$inc_annual[tree.time$yr==2020]*10, na.rm=T)-median(tree.time$inc_annual*10, na.rm=T))/median(tree.time$inc_annual*10, na.rm=T)),2)`% lower than the median across all years. -->
<!-- On average, trees had negative drought sensitivities during drought years, but with considerable variation (Fig `r fig_drought_responses`b).  -->
Median drought sensitivity across all trees in 2010, 2015 and 2020 were `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)`<!-- $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)`-->, `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)`<!-- $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`--> and `r round(median(tree.time$sens.prop[tree.time$yr==2020]), 2)`<!-- $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2020]), 2)`--> respectively. 
<!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->

<!-- Drought years were associated with lower growth across the plot.  -->
<!-- Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4).  -->
<!-- Across the years, on average, `r round(100*mean(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% of the trees had negative growth, with a minimum of `r round(100*min(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))),2)`% in `r names(sort(tapply(tree.time$inc_annual, tree.time$yr, function(x) sum(x < 0) / length(x))))[1]`.  -->
<!-- The drought year 2015 had the maximum proportion of trees with negative growth in the timeseries, `r round(100*sum(tree.time$inc_annual[tree.time$yr==2015]<0)/nrow(tree.time[tree.time$yr==2015,]),2)`%. -->
<!-- *(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))* -->

@@

Species median sensitivities and individual tree sensitivities to the three droughts were not correlated with each other, supporting prediction (*i*) (Fig S6, Fig S7). 
Median annual increment across years varied among the `r nrow(tree.time.sp)` species analysed, with *`r tree.time.sp$spfull[1]`* the lowest (`r tree.time.sp$inc_annual_med[1]`<!-- $\pm$ `r tree.time.sp$inc_annual_sd[1]`--> mm) and *`r tree.time.sp$spfull[nrow(tree.time.sp)]`* the highest (`r tree.time.sp$inc_annual_med[nrow(tree.time.sp)]`<!-- $\pm$ `r tree.time.sp$inc_annual_sd[nrow(tree.time.sp)]`--> mm).
In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/3]`*. 
In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/3]`*, and in 2020 from `r sens.sp[sens.sp$yr==2020,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2020,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[nrow(sens.sp)/3]`*. 
Species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees <!--(for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020, r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for 2010 and 2020, r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, -->(Fig S6). Median species sensitivities to the 2010 and 2020 droughts were variable, but no species had higher median increments than a typical year in 2015. <!--KAT reworded here because Mitrephora thorelii was very slightly positive-->

```{r model summaries, echo=F, warning=F, message=F}
# load data--------------------------
datasets <- readRDS("../data/HKK-dendro/sensitivity_data.RData")

tree.time <- datasets$tree.time
tree_vars <- datasets$tree_vars
sp_vars <- datasets$sp_vars

colnames(sp_vars)[1] <- "Species"

>>>>

@@

# Data Accessibility Statement {-}
<!-- This manuscript does not use novel code.  -->
<!-- The dendrometer band data used for this manuscript is archived in the ForestGEO data repository and will be made available upon request.  -->
<!-- The climate data used in this work are publicly available.  -->
Data and code required to reproduce the results and figures are archived in Zenodo with [doi:10.5281/zenodo.15777974](https://zenodo.org/records/15777975).

@@

```{r data summaries, echo=F, warning=F, message=F}
library(tidyverse)
library(here)
tree.time <- read.csv(here("data", "dendro", "sensitivity_dataset.csv"))
ann.summary <- read.csv(here("data", "dendro", "summaries_dataset.csv"))
# tree.time <- read.csv("data/dendro/sensitivity_dataset.csv")
# ann.summary <- read.csv("data/dendro/summaries_dataset.csv")

sens.sp <- tree.time %>%
  group_by(spfull, yr) %>%
  dplyr::summarise(
    sens_med = round(median(sens.prop, na.rm = T), 2),
    sens_mean = round(mean(sens.prop, na.rm = T), 2),
    sens_sd = round(sd(sens.prop, na.rm = T), 2)
  ) %>%
  arrange(sens_med)

@@

Drought years were associated with lower plot-level growth and negative tree-level sensitivities, with the strongest negative response in 2015 (Fig `r fig_drought_responses`). 
Median growth rate across all trees and years was `r round(ann.summary$median_inc_mm[ann.summary$yr=="all"],2)` mm (range -14.2, 29.32). 
The drought year 2015 had the lowest median increment `r round(ann.summary$median_inc_mm[ann.summary$yr=="2015"],2)` mm, `r round(-100*((ann.summary$median_inc_mm[ann.summary$yr=="2015"]-ann.summary$median_inc_mm[ann.summary$yr=="all"])/ann.summary$median_inc_mm[ann.summary$yr=="all"]),0)`% lower than the median across all years. 
In the 2010 and 2020 droughts, the median increments were `r round(ann.summary$median_inc_mm[ann.summary$yr=="2010"],2)` mm and `r round(ann.summary$median_inc_mm[ann.summary$yr=="2020"],2)` mm respectively, `r round(-100*((ann.summary$median_inc_mm[ann.summary$yr=="2010"]-ann.summary$median_inc_mm[ann.summary$yr=="all"])/ann.summary$median_inc_mm[ann.summary$yr=="all"]),0)`% and `r round(-100*((ann.summary$median_inc_mm[ann.summary$yr=="2020"]-ann.summary$median_inc_mm[ann.summary$yr=="all"])/ann.summary$median_inc_mm[ann.summary$yr=="all"]),0)`% lower than the median across all years. 
Median drought sensitivity across all trees in 2010, 2015 and 2020 were `r round(median(tree.time$sens.prop[tree.time$yr==2010]), 2)`<!-- $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2010]), 2)`-->, `r round(median(tree.time$sens.prop[tree.time$yr==2015]), 2)`<!-- $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2015]), 2)`--> and `r round(median(tree.time$sens.prop[tree.time$yr==2020]), 2)`<!-- $\pm$ `r round(sd(tree.time$sens.prop[tree.time$yr==2020]), 2)`--> respectively. 
<!-- *(See [issue #27](https://github.com/forestgeo/growth-precip-thailand/issues/27))* -->

<!-- Drought years were associated with lower growth across the plot.  -->
<!-- Besides decrease in magnitude of growth, the proportion of trees with negative growth was also higher in the drought years (Fig S4).  -->
<!-- *(See [issue #28](https://github.com/forestgeo/growth-precip-thailand/issues/28))* -->

@@

Species median sensitivities and individual tree sensitivities to the three droughts were not correlated with each other, supporting prediction (*i*) (Fig S6, Fig S7). 
Median annual increment across years varied among the 30 species analysed, with *Gluta obovata* the lowest (0.61 mm) and *Dipterocarpus alatus* the highest (3.58 mm).
In 2010, species-level sensitivities ranged from `r sens.sp[sens.sp$yr==2010,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2010,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2010,]$spfull[nrow(sens.sp)/3]`*. 
In 2015, these ranged from `r sens.sp[sens.sp$yr==2015,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2015,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2015,]$spfull[nrow(sens.sp)/3]`*, and in 2020 from `r sens.sp[sens.sp$yr==2020,]$sens_med[1]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[1]`* to `r sens.sp[sens.sp$yr==2020,]$sens_med[nrow(sens.sp)/3]` for *`r sens.sp[sens.sp$yr==2020,]$spfull[nrow(sens.sp)/3]`*. 
Species had more negative sensitivities to the 2015 drought than the 2010 or 2020 droughts, in alignment with median sensitivities across all trees <!--(for 2010 and 2015 Pearson's correlation coefficient r = `r round(sens.sp.cor_10_15[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_15[3]$p.value,2)`, for 2015 and 2020, r = `r round(sens.sp.cor_15_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_15_20[3]$p.value,2)`, and for 2010 and 2020, r = `r round(sens.sp.cor_10_20[4]$estimate, 2)`, p = `r round(sens.sp.cor_10_20[3]$p.value,2)`, -->(Fig S6). Median species sensitivities to the 2010 and 2020 droughts were variable, but no species had higher median increments than a typical year in 2015. <!--KAT reworded here because Mitrephora thorelii was very slightly positive-->

```{r model summaries, echo=F, warning=F, message=F}
# make sp_vars data
sp_vars <- tree.time %>%
  group_by(Species, spfull) %>%
  dplyr::summarise(williams_dec = first(williams_dec))
